---
title: "Final project"
author: "Abdulaziz Bokijonov"
output: html_notebook
---

## Research Question 

My research data is student alcohol consumption. This dataset has columns indicating school, sex, age, family size, parents cohabitation status, mothers and fathers education, travel time from home to school, and number of past classes failed and etc. for all cases, which are students.

In addition to this data set, there is one more about life expectancy. Cases are countries from 2000 till 2015 followed by data about life expectancy, alcohol, deaths and etc.

Using this data I want to find if there is any correlation between those variables and the alcohol consumption of students. Which factors influence the alcoholism among students the most, does alcohol influence the school work, do some combination of variables cause more alcoholism and does it change over time? Does alcohol effects mortality? Any other interesting trends and correlations will be addressed as well. 

It is an interesting research, because as a college student I am eager to learn the insights of my own college life and what factors influence it and what are results I might expect from it, although the data is from Portuguese institution. Also, from the second data set, I want to see how badly does alcohol affect life expectancy and overall health. Moreover, this research can prevent death or sicknesses from intoxication after finding out the causes of alcohol consumption.



## Data Processing 

#### Data Intake 

```{r}
library(tidyverse)
library(dcData)
DataTable <- read.csv("student-alcohol-consumption.csv")
DataTableLife <- read.csv("Life Expectancy Data.csv")
head(DataTable)
head(DataTableLife)
```

To use the data, I just downloaded the data csv files from Kaggle website and used read.csv function to import them into the r environment.

In the first dataset, the cases are student and here are the key variables:

* school: The educational institution the student attends (categorical: "GP" - Gabriel Pereira, "MS" - Mousinho da Silveira).
* sex: The gender of the student (categorical: "F" - female, "M" - male).
* age: The age of the student (quantitative: 15-22 years old).
* famsize: The size of the student's family (categorical: "GT3" - more than 3, "LE3" - 3 or fewer members).
* Pstatus: The living arrangement of the student's parents (categorical: "T" - living together, "A" - living apart).
* Medu: The educational level of the student's mother (quantitative: 0-4, representing different levels of education).
* Fedu: The educational level of the student's father (quantitative: 0-4, representing different levels of education).
* traveltime: The duration of travel from home to school (quantitative: 1-4, representing different time levels).
* failures: The number of previous class failures (quantitative: 0-4).
* Dalc: Alcohol consumption on workdays (quantitative: 1-5, ranging from very low to very high).
* Walc: Alcohol consumption on weekends (quantitative: 1-5, ranging from very low to very high).
* health: The student's current health status (quantitative: 1-5, ranging from very bad to very good).
* absences: The number of school absences (quantitative: 0-93).
* G3: The final grade achieved by the student (quantitative: 0-20).

Potential outliers of this data can be people who are acing both alcoholism and academics, because I suppose alcoholism would be negatively reflected on the academics

In the second data set, cases are countries from 2000 till 2015 followed by these key variables:

* country (categorical)
* year (quantitative: year of data collection)
* Life expectancy (quantitative),
* Alcohol: Quantitative variable representing the recorded per capita (15+) consumption of alcohol in litres of pure alcohol.
* Adult Mortality Rates: Quantitative variable representing the probability of dying between 15 and 60 years per 1000 population for both sexes.

The ourliers for this data set can be countries that don't have high quality of life, thus more mortality rate not related to alcohol.

#### Data Wrangling 

```{r}

# Find the most popular value in the column
MostPopularVal <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Groups the data frame by totalAlc and find mean for numbers and most popular value for characters
GroupEverything <- function(dataSet) {
  ToReturn <- dataSet %>%
    group_by(totalAlc) %>%
    summarise(school = MostPopularVal(school))
  for (i in 2:(ncol(dataSet)-1)){
     varType <- class(dataSet[,i])
     if(as.character(varType) == "character"){
      tempData <- dataSet %>%
        group_by(totalAlc) %>%
        summarise(
          MostPopularVal(eval(parse(text = paste(names(dataSet)[i]))))
        )
     }
     else{
      tempData <- dataSet %>%
        group_by(totalAlc) %>%
        summarise(
          mean(eval(parse(text = paste(names(dataSet)[i]))))
        ) 
     }
     
     ToReturn <- merge(ToReturn, tempData)
     names(ToReturn)[length(names(ToReturn))] <- names(dataSet)[i]
     
  }
  return(ToReturn)
}

# Adds workday and weekend alcohol using mutate
AlcoholTotal <-
  DataTable %>%
  mutate(totalAlc = Dalc+Walc) 

# Gets all grouped table
StudentsGroupedTotalAlc <-GroupEverything(AlcoholTotal)
StudentsGroupedTotalAlc
# Alcohol consumption and failed classes count
AlcoholFClasses <-
  DataTable %>%
  group_by(failures) %>%
  summarise(meanAlcWeekday = mean(Dalc),
            meanAlcWeekend = mean(Walc))
head(AlcoholFClasses)

# Alcohol consumption and age
AlcoholAge <-
  DataTable %>%
  group_by(age) %>%
  summarise(meanAlcWeekday = mean(Dalc),
            meanAlcWeekend = mean(Walc))
head(AlcoholAge)

# Alcohol consumption and life expectancy with deaths
AlcoholLife <- DataTableLife %>%
  group_by(Alcohol) %>%
  filter(Status == "Developed")
head(AlcoholLife)
  
``` 




## Data Visualization

```{r}
AlcoholFClasses %>%
  ggplot(aes(x=failures,y=(meanAlcWeekday+meanAlcWeekend))) +
  geom_bar(stat='identity',position='stack', width=.9) +
  labs(y= "Total alcohol week", x = "Failed classes") + 
  ggtitle("Alcohol consumed during the week and failed classes")

StudentsGroupedTotalAlc %>%
  ggplot(aes(x=totalAlc,y=freetime, color=sex)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  labs(y= "Free time", x = "Total Alcohol") + 
  ggtitle("Alcohol consumed and Free time")

StudentsGroupedTotalAlc %>%
  ggplot(aes(x=totalAlc,y=G3, color=sex)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  labs(y= "Final grade", x = "Total Alcohol") + 
  ggtitle("Alcohol consumed and Final grades")
```
* In the first graph we see that the there is positive correlation between alcohol consumed and failed classes. This graph shows that more alcohol consumed, more classes failed. 

* In the second graph, we can see the relation of free time and alcohol consumed. Both for females and males, more free time correlates with more alcohol consumed. This can be both cause and effect of alacohol consumption.

* In the third graph we can see the final grades relation to the total alcohol consumed. For females alcohol consumption negatively relates to grades, meaning that the more alcohol consumed, the lower the final grades. For males its the opposite, alcohol consumtion have positive relation with their final grades.

```{r}
AlcoholLife %>%
  ggplot(aes(x=Alcohol,y=Life.expectancy, color=Year)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(y= "Life expectancy", x = "Alcohol") + 
  ggtitle("Alcohol consumed and life expectancy")

AlcoholLife %>%
  ggplot(aes(x=Alcohol,y=BMI, color=Year)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(y= "BMI", x = "Life expectancy") + 
  ggtitle("Alcohol consumed and BMI")
```

The second dataset's countries were filtered to only Developed contries to avoid outliers.

* In the first graph, based on the best fit line, there is a negative relation between alcohol consumed and life expectancy. More alcohol consumed, lower the life expectancy.

* In the second graph, BMI (body mass index) positively relates to alcohol consumed. Its higher when more alcohol is consumed. P.S., high can cause health problems and decrease life expectancy.

```{r}
AlcoholAge %>%
  ggplot(aes(x=age,y=(meanAlcWeekday+meanAlcWeekend))) +
  geom_bar(stat='identity',position='stack', width=.9) +
  labs(y= "Total alcohol week", x = "Age") + 
  ggtitle("Alcohol consumed during the week by age")

StudentsGroupedTotalAlc %>%
  ggplot(aes(y=totalAlc,x=traveltime, color=sex)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  labs(x= "Travel time from home", y = "Total Alcohol") + 
  ggtitle("Alcohol consumed and Travel time from home")

StudentsGroupedTotalAlc %>%
  ggplot(aes(y=totalAlc,x=health, color=sex)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  labs(x= "Health", y = "Total Alcohol") + 
  ggtitle("Alcohol consumed and Final grades")

StudentsGroupedTotalAlc %>%
  ggplot(aes(y=totalAlc,x=famrel, color=sex)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  labs(x= "Family Relation", y = "Total Alcohol") + 
  ggtitle("Alcohol consumed and Family Relation")
``` 
In this section we are trying to indentify the "cause" of alcohol consumtion, at lest the variables that have positive relation with it.

* First graph shows the bar chart of alcohol consumtion by age. Age has positive relation with alcohol consumption, the older the student, the more they drink

* Second graph demonstrates the positive relation between travel time from home and alcohol consumtion for both males and females, but the slope for females is bigger. This means that the further is the home, more alcohol is consumed by the student.

* Third graph demonstrates the positive relation between health and alcohol consumtion for both males and females, but the slope for females is bigger.

* The last graph shows the relation between family relation (from 0 /bad to 5/good ) and there is the small positive relation from males and significant negative relation for females. Females drink less when family relations worsen (in this dataset)



## Conclusion

  Research question were to try to indentify what causes alcohol consumtion among students and with it cases when it comes to studies. The research above showed that the there is positive correlation between alcohol consumed and failed classes. In addition to that we saw that the final grades relation to the total alcohol consumed for females was negative, meaning that the more alcohol consumed, the lower the final grades. For males was the opposite, alcohol consumption had positive relation with their final grades. One more interesting finding was that both males and females had more free time when consuming more alcohol, or it could be visa versa, consuming more alcohol when having more free time. Thus, overall, higher consumtion of alcohol relates to more failed classes, lower grades for females, higher grades for males, and more free time for both sex.
    When it comes to trying to indetify the "causes" of alcohol consumtion, we saw a very string positive relation with age and alcohol. People drink more when the get older, according to the graph. Also the alcohol consumtion increased when travel time of the student from home increased. Similar relation was detected with health, the better the health condition, more alcohol is consumed.
    One more finding was that the life expectancy in developed contries decreased and BMI increased as more alcohol was consumed. It can indicate that alcohol worsens the health conditions.
    To sum up, all research question were fairly answered and data wrangling and visualization were done in the most accurate way possible.
